---
title: "The impact of firebreaks in Wales and Northern Ireland on reproduction numbers"
author: "Kath Sherratt, Sam Abbott and Sebastian Funk on behalf of the CMMID Covid-19 Working Group"
date: "17/11/2020"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy = TRUE, tidy.opts = list(comment = FALSE))
```

## Rationale

We aimed to explore ongoing intensive lockdowns (firebreaks) in Northern Ireland (from 16th October 2020) and Wales (24th October to 9th November 2020), by analysing the reproduction number (R~t~) before and after the intervention. In order to identify any effect from these interventions, we estimated R~t~ with set breakpoints on the dates the interventions came into place.

## Methods

We estimated R~t~ separately from cases, hospitalisations and deaths (shown as a 7 day moving average in Fig 1A) from Monday 28th September using two different methods.

First we modelled R~t~ as piecewise constant with a single breakpoint at the start of respective intervention (figure 2B). Second, we modelled R~t~ using a weekly random walk  with an additional breakpoint at the start of respective lockdowns. In the latter method, we left at least one week between the last random step and the firebreak breakpoint. No further breakpoints were included beyond the start of the firebreak until the end of the firebreak (9 November in Wales). In all estimates, we specified an R~t~ prior of mean 1.2 (SD 0.2). 

We measured effect size as a multiplicative change in R~t~.

## Results

We saw some evidence of an impact of the firebreak when R~t~ was modelled using breakpoint. The firebreak appeared to have a strong effect on decreasing R~t~ by ca. 10-30% when estimated from hospital admissions in Northern Ireland and Wales, bringing R~t~ below 1 in both cases. A similar effect was observed in Rt from reported cases in Wales.

```{r, fig.height = 2, fig.width = 7, fig.align='center', message=F, warning=F, echo=FALSE}
 library(dplyr, quietly = TRUE); library(ggplot2, quietly = TRUE)

effects <- suppressMessages(readr::read_csv(here::here(Sys.Date(),
                            "breakpoint-effects.csv"))) %>%
  dplyr::mutate_if(is.numeric, ~ 1 - .) 

effects %>%
  ggplot() +
  geom_point(aes(x = median, y = region, colour = model),
             position = position_dodge(width = 0.2)) +
  geom_linerange(aes(xmin = lower_90, xmax = upper_90,
                     y = region, colour = model),
                 alpha = 0.3, position = position_dodge(width = 0.2)) +
  geom_linerange(aes(xmin = lower_50, xmax = upper_50,
                     y = region, colour = model),
                 alpha = 1, position = position_dodge(width = 0.2)) +
  geom_vline(aes(xintercept = 0), lty = 3) +
  facet_wrap(~source) +
  labs(y = NULL, x = "Effect of intervention on Rt") +
  scale_x_continuous(labels = scales::label_percent(accuracy = 1L)) +
  scale_colour_brewer("Model", type = "qual", palette = 2) +
  theme_classic() +
  theme(legend.position = "bottom",
        strip.background = element_rect(colour = "transparent"))
```
*Figure 1. Effect size (multiplicative change shown as percentage) of breakpoint on R~t~, on the start of the firebreak: 24th October for Wales, 16th October for N Ireland, median with 50% (darker) and 90% (lighter) credible intervals.*


```{r, fig.height = 12, fig.width = 15, message=F, warning=F, echo=FALSE}
library(patchwork, quietly = TRUE)
# Set up plot data --------------------------------------------------------------
# Get formatted data
data_list <- readRDS(here::here(Sys.Date(), 
                              "formatted-data.rds"))
data <- data_list$data
data_ma <- data_list$data_ma

# Get model estimates
models <- suppressMessages(readr::read_csv(here::here(Sys.Date(), 
                                     "firebreak-breakpoints.csv")))

# Set fixed values
regions <- unique(as.character(data$region))
break_ni_start <- as.Date("2020-10-16")
break_ni_end <- as.Date("2020-11-20")
break_wales_start <- as.Date("2020-10-24")
break_wales_end <- as.Date("2020-11-09")

colours <- c("cases" = "#1b9e77",  "admissions" =  "#7570b3", "deaths" = "#d95f02")

# Plot data --------------------------------------------------------------------
plot_data_fn <- function(region_name, 
                         breakpoint_start = NA, breakpoint_end = NA,
                         value_type, title = FALSE){
  data_ma %>%
    dplyr::mutate(source = factor(source)) %>%
    dplyr::filter(region %in% region_name &
                  date >= min(models$date),
                  source == value_type) %>%
    ggplot() +
    geom_line(aes(x = date, y = (as.numeric(ma)),
                  colour = source)) +
    geom_vline(xintercept = as.Date(breakpoint_start),
               lty = 3, colour = "grey50") +
    geom_vline(xintercept = as.Date(breakpoint_end),
               lty = 3, colour = "grey50") +
    geom_point(data = data %>%
               dplyr::mutate(source = factor(source)) %>%
               dplyr::filter(region %in% region_name,
                             date >= min(models$date),
                             source == value_type),
               aes(x = date, y = value, colour = source)) +
    cowplot::theme_cowplot() +
    # coord_cartesian(xlim = c(date_min, date_max)) +
    scale_color_manual("", values = colours, drop = FALSE) +
    scale_x_date(date_breaks = "1 week", date_labels = "%b %d") +
    theme(axis.text.x = ggplot2::element_text(angle = 30, hjust = 1)) +
    guides(colour = FALSE) +
      labs(y = value_type, x = NULL) +
      scale_y_log10() -> p
  if (title) {
    p <- p + ggtitle(region_name)
  }
  return(p)
}

# Plot Rt  ----------------------------------------------------------
plot_rt_fn <- function(region_name, model_name, 
                       breakpoint_start = NA,
                       breakpoint_end = NA){
  models %>%
    dplyr::filter(region %in% region_name & 
                    model %in% model_name) %>%
    ggplot2::ggplot(ggplot2::aes(x = date, col = source, fill = source)) +
    ggplot2::geom_ribbon(ggplot2::aes(ymin = lower_50, ymax = upper_50),
                         alpha = 0.2, size = 0, colour = NA) +
    ggplot2::geom_ribbon(ggplot2::aes(ymin = lower_90, ymax = upper_90),
                         alpha = 0.1, colour = NA) +
    ggplot2::geom_line(ggplot2::aes(y = median),
                       alpha = 0.9, size = 1) +
    ggplot2::geom_vline(xintercept = as.Date(breakpoint_start), 
                        lty = 3, colour = "grey50") +
    ggplot2::geom_vline(xintercept = as.Date(breakpoint_end), 
                        lty = 3, colour = "grey50") +
    ggplot2::geom_hline(yintercept = 1, linetype = 2) +
    ggplot2::scale_color_manual("", values = colours) +
    ggplot2::scale_fill_manual("", values = colours) +
    ggplot2::scale_x_date(date_breaks = "1 week", date_labels = "%b %d") +
    #scale_y_continuous(breaks=seq(0, 1.4, by = 0.2)) +
    cowplot::theme_cowplot() +
    ggplot2::labs(y = paste("R(t)", model_name), x = NULL) +
    ggplot2::theme(legend.position = "bottom",
                   axis.text.x = ggplot2::element_text(angle = 30, hjust = 1)) +
    ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(alpha = 1)))
}

# Combine plots -----------------------------------------------------------
# Plot Rt
# Wales
bp_admissions_wales <- plot_rt_fn(region_name = "Wales", 
                               model_name = "single breakpoint",
                                breakpoint_start = break_wales_start,
                               breakpoint_end = break_wales_end)
rw_admissions_wales <- plot_rt_fn(region_name = "Wales", 
                               model_name = "random walk + breakpoint",
                                breakpoint_start = break_wales_start,
                               breakpoint_end = break_wales_end)
# NI
bp_admissions_ni <- plot_rt_fn(region_name = "Northern Ireland", 
                               model_name = "single breakpoint",
                                breakpoint_start = break_ni_start,
                               breakpoint_end = break_ni_end)
rw_admissions_ni <- plot_rt_fn(region_name = "Northern Ireland", 
                               model_name = "random walk + breakpoint",
                                breakpoint_start = break_ni_start,
                               breakpoint_end = break_ni_end)

# Plot data
# Wales
admissions_wales <- plot_data_fn(region_name = "Wales", 
                                 breakpoint_start = break_wales_start,
                               breakpoint_end = break_wales_end,
                                 value_type = "admissions", title = TRUE)
cases_wales <- plot_data_fn(region_name = "Wales", 
                             breakpoint_start = break_wales_start,
                               breakpoint_end = break_wales_end,
                            value_type = "cases")
deaths_wales <- plot_data_fn(region_name = "Wales", 
                              breakpoint_start = break_wales_start,
                               breakpoint_end = break_wales_end,
                             value_type = "deaths")
# NI
admissions_ni <- plot_data_fn(region_name = "Northern Ireland", 
                              breakpoint_start = break_ni_start,
                               breakpoint_end = break_ni_end, 
                              value_type = "admissions", title = TRUE)
cases_ni <- plot_data_fn(region_name = "Northern Ireland", 
                         breakpoint_start = break_ni_start,
                               breakpoint_end = break_ni_end, 
                         value_type = "cases")
deaths_ni <- plot_data_fn(region_name = "Northern Ireland", 
                          breakpoint_start = break_ni_start,
                               breakpoint_end = break_ni_end, 
                          value_type = "deaths")

# Join plots
plot_breaks <- suppressWarnings(((admissions_wales | admissions_ni) +
                  plot_annotation(subtitle = "A. Admissions")) /
  ((cases_wales | cases_ni) +
   plot_annotation(subtitle = "B. Cases")) /
  ((deaths_wales | deaths_ni) +
   plot_annotation(subtitle = "C. Deaths")) /
  ((bp_admissions_wales | bp_admissions_ni) +
     plot_annotation(subtitle = "D. Single breakpoint")) /
  ((rw_admissions_wales | rw_admissions_ni) +
     plot_annotation(subtitle = "E. Random walk with breakpoint")) +
  plot_layout(guides = "collect")  &
  theme(legend.position = "bottom"))

plot_breaks

```
*Figure 2. Top to bottom: Data for hospital admissions, cases, and deaths, in Wales and Northern Ireland. R~t~ modelled using a single breakpoint: 24th October in Wales (firebreak); 16th October in Northern Ireland (firebreak). R~t~ including a random walk, with breakpoints each Sunday until the week before the firebreak, plus the single breakpoint as above.*
